---
title: "DE"
author: "Howard Cen"
date: "2024-10-01"
output: html_document
---


```{r }
library(tidyverse)
library(readxl)



if (!require("protti", quietly = TRUE))
    devtools::install_github("jpquast/protti", dependencies = TRUE)

if (!requireNamespace("tidyverse", quietly = TRUE))
    install.packages("tidyverse")
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
library(protti) # Main proteomics analysis package
# Load functions from utils.R
source("utils.R")
```

```{r adjust pval}
perseus.df <- read_excel("input/df.summary_proteomics.xlsx")

perseus.df$padj <- p.adjust(perseus.df$pval, method = "BH")

View(perseus.df)

write.csv(perseus.df, "output/GFP_proteomics_perseus_DE_results_high_low.csv", row.names = F)
```

```{r protti}

pro.anno <- read_excel("input/raw values.xlsx")
View(pro.anno)
pro <- read.delim("input/Mouse raw data non normalized.tsv")
pro <- read.delim("input/Mouse raw data normalized.tsv")

View(pro)
colnames(pro)

colnames(pro)[6:38] <- stringr::str_extract(colnames(pro)[6:38], "[A-Z0-9]+_[A-Za-z]+GFP")
length(unique(pro$Protein.Group))==length(pro$Protein.Group)
length(unique(pro$Genes))==length(pro$Genes)

# remove F1_Low, F1_High, F4_High, F6_Low, F6_High, M4_High
# remove F1_Neg, M4_Neg, M4_Low
# remove F5, F6_Neg
# remove F4
# remove M5
protein_data <- pro[c(grep("Group", colnames(pro)), grep("Neg", colnames(pro)), grep("Low", colnames(pro)), grep("High", colnames(pro)))]

protein_data <- protein_data[-grep("F1_Low|F1_High|F4_High|F6_Low|F6_High|M4_High", colnames(protein_data))]

protein_data <- protein_data[-grep("F1_Neg|M4_Neg|M4_Low", colnames(protein_data))]

protein_data <- protein_data[-grep("F5|F6|F4", colnames(protein_data))]

protein_data <- protein_data[-grep("M5", colnames(protein_data))]

# protein_data <- protein_data[-grep("Neg", colnames(protein_data))] # removing Neg sample made the stats worse

# kept proteins that are detected in at least 1/3 samples
nrow(protein_data)
protein_data <- protein_data[rowSums(is.na(protein_data[-1]))<ncol(protein_data[-1])*2/3,]
  
#  pro.dat1 <- protein_data[,grep("Group|High", colnames(protein_data))]
#  pro.dat1 <- pro.dat1[rowSums(!is.na(pro.dat1))>=3,]
  
#  pro.dat2 <- protein_data[,grep("Group|Low", colnames(protein_data))]
#  pro.dat2 <- pro.dat2[rowSums(!is.na(pro.dat2))>=3,]
  
#  id.filt <- intersect(pro.dat1$Protein.Group, pro.dat2$Protein.Group)
  
#protein_data <- protein_data %>% filter(Protein.Group %in% id.filt)

nrow(protein_data)

#protein_data <- protein_data[rowSums(is.na(protein_data[-1]))==0,]
# 5940 --> 5298

# meta data

metadata <- data.frame(ID = colnames(protein_data)[-1]) %>%
  mutate(Group = gsub("[A-Z][0-9]_", "", ID)) %>%
  mutate(Group2 = ifelse(Group == "NegGFP", "NegGFP", "PosGFP"))
View(metadata)

```


```{r}
# Convert the peptide_data to long format
protein_data_long <- tidyr::pivot_longer(
    protein_data,
    cols = -c("Protein.Group"),
    names_to = "ID",
    values_to = "Intensity"
)
head(protein_data_long)

# Merge the metadata and peptide_data_longer
protein_data_long <- dplyr::left_join(
    protein_data_long,
    metadata,
    by = "ID"
)

# Rename the columns
protein_data_long <- dplyr::rename(
    protein_data_long,
    "Protein" = "Protein.Group",
    "Sample" = "ID"
)

# Create a log2 transformed intensity column
protein_data_long$Intensity_log2 <- log2(protein_data_long$Intensity)


```
# BONUS 1 - Synthetic Data Generation
# Creates 100 Proteins and their replicates with 5 replicates over 2 conditions
#data <- protti::create_synthetic_data(
#    n_proteins = 100,       # number of proteins
#    frac_change = 0.05,
#    n_replicates = 5,
#    n_conditions = 2,
#    method = "effect_random",
#    additional_metadata = TRUE
#)

```{r}
## Quality Checks and Data Filtering

# 1. Number of Identified Proteins per Samples
plot <- qc_ids(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    condition = Group,
    intensity = Intensity,
    plot = TRUE
)

# Visualize the plot

# Sort the bars by the number of proteins
plot + 
  ggplot2::coord_flip() + 
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 0, hjust = 0.5))


# Rotate the x-axis labels
plot + ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5)
)

table(protein_data_long[!is.na(protein_data_long$Intensity),]$Sample)
```
```{r}
pdf("IDcount.pdf", width = 8, height = 4)
# Rotate the x-axis labels
plot + ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5)
)
dev.off()
```



```{r}

# 2. Intensity Distribution and Median Intensity Plots
#   a. Intensity Distribution
qc_intensity_distribution(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity_log2 = Intensity,
    plot_style = "boxplot"
)

# Plot again with log2 transformed intensity
qc_intensity_distribution(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity_log2 = Intensity_log2,
    plot_style = "boxplot"
) + ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5))
```
```{r}
#   b. Median Intensity Plot
qc_median_intensities(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity = Intensity_log2
) + ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 90, hjust = 1, vjust = 0.5))
```

```{r}

# 3. Co-efficients of Variation (CV)
# Within Sample CVs - Usually 10 or 20% are good
qc_cvs(
    data = protein_data_long,
    grouping = Protein,
    condition = Group,
    intensity = Intensity,
    plot = TRUE,
    plot_style = "boxplot"
    #max_cv = 1000
)

```

```{r}
# 4. Data Completeness -  what percentage of all detected precursors is present in each sample.
qc_data_completeness(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity = Intensity_log2,
    plot = TRUE
)

```


```{r}
# 5. Sample Correlation
# NOTE: Requires pheatmap & seriation packages
#install.packages('seriation')

qc_sample_correlation(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity = Intensity_log2,
    condition = Group,
    interactive = FALSE,
    method = "pearson"
)
# can't display the heatmap when it's ran inside code chunk. Works outside below


```


qc_sample_correlation(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity = Intensity_log2,
    condition = Group,
    interactive = FALSE,
    method = "pearson"
)

```{r}
# 6. Principal Component Analysis (PCA)

qc_pca(
  data = protein_data_long,
  sample = Sample,
  grouping = Protein,
  intensity = Intensity_log2,
  condition = Group,
  digestion = NULL,
  plot_style = "scree"
)
```

```{r}
qc_pca(
  data = protein_data_long,
  sample = Sample,
  grouping = Protein,
  intensity = Intensity_log2,
  condition = Group,
  digestion = NULL,
  plot_style = "pca"
)
```

```{r}
# 7. Ranked Intensity Distribution (Protein-Rank Plot)
qc_ranked_intensities(
  data = protein_data_long,
  sample = Sample,
  grouping = Protein,
  intensity_log2 = Intensity_log2,
  plot = TRUE,
  y_axis_transformation = "log2"
)

```

```{r, include=FALSE}
# 8. Removing Problematic Samples
# Remove instances where SampleName == "NP21"
#protein_data_long <- dplyr::filter(
#    protein_data_long,
#    SampleName != "NP21"
#)

# 9. Removing Highly Missing Proteins
# Remove proteins with missing values in more than 75% of samples
# TODO: This is a custom solution, WIP


#colnames(protein_data_long)
#nsample <- length(unique(protein_data_long$Sample))
#protein_data_long_filter <- protein_data_long %>%
#  group_by(Protein) %>%
#  filter(sum(is.na(Intensity_log2))<nsample*0.75)

```



```{r}
## Data Normalization

# Median Normalization using protti
protein_data_long <- normalise(
    data = protein_data_long,
    sample = Sample,
    intensity = Intensity_log2,
    method = "median"
) # Adds normalized_"intensity" column

```

```{r}
b1 <- qc_median_intensities(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity = Intensity_log2
) + ggplot2::ggtitle("Before Normalization")

b2 <- qc_median_intensities(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity = normalised_intensity_log2
) + ggplot2::ggtitle("After Normalization")

# Plot the two plots side-by-side

cowplot::plot_grid(
    b1, b2,
    ncol = 2
)

```
```{r}
qc_pca(
  data = protein_data_long,
  sample = Sample,
  grouping = Protein,
  intensity = normalised_intensity_log2,
  condition = Group,
  digestion = NULL,
  plot_style = "pca"
)

```

```{r save PCA}
pdf("PCA_norm.pdf", width = 6.5, height = 5)
qc_pca(
  data = protein_data_long,
  sample = Sample,
  grouping = Protein,
  intensity = normalised_intensity_log2,
  condition = Group,
  digestion = NULL,
  plot_style = "pca"
)
dev.off()

```

```{r}

pdf("PCA_norm_filter.pdf", width = 6.5, height = 5)
qc_pca(
  data = protein_data_long,
  sample = Sample,
  grouping = Protein,
  intensity = normalised_intensity_log2,
  condition = Group,
  digestion = NULL,
  plot_style = "pca"
)
dev.off()

```

qc_sample_correlation(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity = normalised_intensity_log2,
    condition = Group,
    interactive = FALSE,
    method = "pearson"
)

```{r}
# Imputation of Missing Values

# Impute missing values
protein_data_long <- impute_with_downshifted_normal(
    data = protein_data_long,
    intensity_log2 = normalised_intensity_log2,
    prctl = 0.05,
    downshift_mag = 1.5,
    downshift_min = 0.1
)

p1 <- qc_intensity_distribution(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity_log2 = normalised_intensity_log2,
    plot_style = "violin"
) + ggplot2::theme(
    axis.text.x = ggplot2::element_text(
        angle = 90,
        hjust = 1
    )
) + ggplot2::ggtitle(
    "Non-Imputed Intensity Distribution"
)


p2 <- qc_intensity_distribution(
    data = protein_data_long,
    sample = Sample,
    grouping = Protein,
    intensity_log2 = imputed_intensity_log2,
    plot_style = "violin"
) + ggplot2::theme(
    axis.text.x = ggplot2::element_text(
        angle = 90,
        hjust = 1
    )
) + ggplot2::ggtitle(
    "Imputed Intensity Distribution"
)

cowplot::plot_grid(
    p1, p2,
    ncol = 2,
    align = "v"
)
```

```{r, Limma, include=FALSE}
## Statistical Testing with Weighted Limma package

# Transform the long to wide format
# columns:Sample, rows: Protein, values: imputed_intensity_log2
protein_data_wide <- protein_data_long %>%
    dplyr::select(
        Sample,
        Protein,
        imputed_intensity_log2
    ) %>%
    tidyr::pivot_wider(
        names_from = Sample,
        values_from = imputed_intensity_log2
    ) %>%
    tibble::column_to_rownames(
        var = "Protein"
    )
head(protein_data_wide)

write.csv(protein_data_wide, "output/betaInsGFP_proteomics_log2_norm_impute.csv")


#protein_data_wide <- read.csv("output/liver_proteomics_log2_norm_impute.csv", row.names = 1)

protein_data_wide_non_imputed <- protein_data_long %>%
    dplyr::select(
        Sample,
        Protein,
        normalised_intensity_log2
    ) %>%
    tidyr::pivot_wider(
        names_from = Sample,
        values_from = normalised_intensity_log2
    ) %>%
    tibble::column_to_rownames(
        var = "Protein"
    )

write.csv(protein_data_wide_non_imputed, "output/betaInsGFP_proteomics_log2_norm_nonimpute.csv")


#protein_data_wide_non_imputed <- read.csv("output/liver_proteomics_log2_norm_nonimpute.csv", row.names = 1)


# Find the indices of the missing values
na_index <- which(is.na(protein_data_wide_non_imputed))

# Initialize the weight matrix
weight_matrix <- matrix(
    data = 1,
    nrow = nrow(protein_data_wide),
    ncol = ncol(protein_data_wide)
)
# Weighting of Missing Values
na_weight <- 0.0001
# Replace the missing values with the na_weight
weight_matrix[na_index] <- na_weight

head(weight_matrix)

```

```{r high vs low}
# Create the design matrix

group_levels <- c("NegGFP", "LowGFP", "HighGFP")

#group_levels <- c("LowGFP", "HighGFP")
f <- factor(metadata$Group, levels = group_levels)
design <- model.matrix(~0+f)
colnames(design) <- group_levels
head(design) 

# Variables
pval_thr <- 0.05
log2_fc_thr <- 0

# Fit a linear model with weights
fit <- limma::lmFit(
    protein_data_wide,
    design = design,
    weights = weight_matrix
)


vs <- c(paste0(group_levels[3],"-",group_levels[2]))
#vs <- c(paste0(group_levels[2],"-",group_levels[1]))

contrast_matrix <- limma::makeContrasts(vs[1], #vs[2], vs[3], vs[4], 
                                        levels = design)

fit2 <- limma::contrasts.fit(fit, contrast_matrix)
fit2 <- limma::eBayes(fit2)

library(openxlsx)
#library(org.Hs.eg.db)
library(stringr)

# Create a new Excel workbook
wb <- createWorkbook()
wb.filt <- createWorkbook()
for (i in 1:length(vs)){
  df <- limma::topTable(fit2, n=Inf, coef=i, adjust="BH") %>%
    rownames_to_column(var = "Protein.Group")
  
  anno <- pro %>%
    dplyr::select(any_of(c("Protein.Group", "Protein.Names", "Genes")))
  
  df <- df %>% left_join(anno) %>%
    mutate(Protein.Names = gsub("_MOUSE", "", Protein.Names),
           Genes = gsub("_MOUSE", "", Genes))

    pro.dat1 <- protein_data_wide_non_imputed[,grep(unlist(str_split(vs[i], "-"))[1], colnames(protein_data_wide_non_imputed))]
  pro.dat1 <- pro.dat1[rowSums(!is.na(pro.dat1))>=3,]
  
  pro.dat2 <- protein_data_wide_non_imputed[,grep(unlist(str_split(vs[i], "-"))[2], colnames(protein_data_wide_non_imputed))]
  pro.dat2 <- pro.dat2[rowSums(!is.na(pro.dat2))>=3,]
  id.filt <- intersect(rownames(pro.dat1), rownames(pro.dat2))
  
  df.filt <- df %>% 
    filter(Protein.Group %in% id.filt)
  
  #save to workbook
  addWorksheet(wb, vs[i])
  writeData(wb, vs[i], df)
  
  addWorksheet(wb.filt, vs[i])
  writeData(wb.filt, vs[i], df.filt)
}

saveWorkbook(wb, "output/betaInsGFP_proteomics_limma_DE_results_high_low_15.xlsx", overwrite = TRUE)
saveWorkbook(wb.filt, "output/betaInsGFP_proteomics_limma_DE_results_high_low_filterN3_15.xlsx", overwrite = TRUE)
```

```{r pos vs neg}
# Create the design matrix

group_levels <- c("NegGFP", "PosGFP")
f <- factor(metadata$Group2, levels = group_levels)
design <- model.matrix(~0+f)
colnames(design) <- group_levels
head(design) 

# Variables
pval_thr <- 0.05
log2_fc_thr <- 0

# Fit a linear model with weights
fit <- limma::lmFit(
    protein_data_wide,
    design = design,
    weights = weight_matrix
)


vs <- c(paste0(group_levels[2],"-",group_levels[1]))

contrast_matrix <- limma::makeContrasts(vs[1], #vs[2], vs[3], vs[4], 
                                        levels = design)

fit2 <- limma::contrasts.fit(fit, contrast_matrix)
fit2 <- limma::eBayes(fit2)

library(openxlsx)
#library(org.Hs.eg.db)
library(stringr)

# Create a new Excel workbook
wb <- createWorkbook()
wb.filt <- createWorkbook()
{
  df <- limma::topTable(fit2, n=Inf, coef=1, adjust="BH") %>%
    rownames_to_column(var = "Protein.Group")
  
  anno <- pro %>%
    dplyr::select(any_of(c("Protein.Group", "Protein.Names", "Genes")))
  df <- df %>% left_join(anno) %>%
    mutate(Protein.Names = gsub("_MOUSE", "", Protein.Names),
           Genes = gsub("_MOUSE", "", Genes))
  
  
  pro.dat1 <- protein_data_wide_non_imputed[,grep("Neg", colnames(protein_data_wide_non_imputed))]
  pro.dat1 <- pro.dat1[rowSums(!is.na(pro.dat1))>=3,]
  
  pro.dat2 <- protein_data_wide_non_imputed[,grep("High|Low", colnames(protein_data_wide_non_imputed))]
  pro.dat2 <- pro.dat2[rowSums(!is.na(pro.dat2))>=3,]
  id.filt <- intersect(rownames(pro.dat1), rownames(pro.dat2))
  
  df.filt <- df %>% 
    filter(Protein.Group %in% id.filt)
  
  #save to workbook
  addWorksheet(wb, vs[i])
  writeData(wb, vs[i], df)
  
  addWorksheet(wb.filt, vs[i])
  writeData(wb.filt, vs[i], df.filt)
  
}

saveWorkbook(wb, "output/betaInsGFP_proteomics_limma_DE_results_pos_neg.xlsx", overwrite = TRUE)
saveWorkbook(wb.filt, "output/betaInsGFP_proteomics_limma_DE_results_pos_neg_filterN3.xlsx", overwrite = TRUE)

```

```{r}
df <- read_excel("output/betaInsGFP_proteomics_limma_DE_results_high_low_filterN3.xlsx")
protein_data_wide_non_imputed_order <- protein_data_wide_non_imputed[df$Protein.Group,] %>%
  rownames_to_column("Protein.Group") %>%
  left_join(df[c("Protein.Group", "Protein.Names", "Genes")])
View(protein_data_wide_non_imputed_order)
write.csv(protein_data_wide_non_imputed_order, "output/betaInsGFP_proteomics_log2_norm_nonimpute_ordered.csv")

protein_data_wide_non_imputed
```